---
title: "Activity Assignmnet"
author: "Karuna Raghuwanshi"
date: "27/04/2020"
output: html_document
---

```{r, echo=FALSE, results='hide', warning=FALSE, message=FALSE}
library(ggplot2)
library(dplyr)


```

## **Loading and processing the data**
#### 1. Load the data (i.e. read.csv())
```{r, results='markup', warning=TRUE, message=TRUE}
cls = c("integer", "character", "integer")
df <- read.csv("activity.csv", head=TRUE, colClasses=cls, na.strings="NA")
```

#### 2. Process/transform the data (if necessary) into a format suitable for your analysis
```{r}
head(df)
df$date <- as.Date(df$date)
```
----------------------------------------

## **What is mean total number of steps taken per day?** ####
#### 1. Calculate the total number of steps taken per day

```{r}
df_ign <- subset(df, !is.na(df$steps))
dailysum <- tapply(df_ign$steps, df_ign$date, sum, na.rm=TRUE, simplify=T)
dailysum <- dailysum[!is.na(dailysum)]
```

#### 2. Make a histogram of the total number of steps taken each day
```{r}
g <-hist(x=dailysum,
         col="red",
         breaks=20,
         xlab="Daily total steps",
         ylab="Frequency",
         main="The distribution of daily total (missing data ignored)")
```

#### 3. Calculate and report the mean and median total number of steps taken per day
```{r}
mean(dailysum)
median(dailysum)
```
----------------------------------------

## **What is the average daily activity pattern?** ####

#### 1. Make a time series plot  of the 5-minute interval (x-axis) and the 
#### average number of steps taken, averaged across all days (y-axis)
```{r}
int_avg <- tapply(df_ign$steps, df_ign$interval, mean, na.rm=TRUE, simplify=T)
df_ia <- data.frame(interval=as.integer(names(int_avg)), avg=int_avg)

with(df_ia,
     plot(interval,
          avg,
          type="l",
          xlab="5-minute intervals",
          ylab="average steps in the interval across all days"))

```

#### 2. Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?
```{r}
max_steps <- max(df_ia$avg)
df_ia[df_ia$avg == max_steps, ]
```
------------------------------------

## **Imputing missing values**


#### 1. Calculate and report the total number of missing values in the dataset (i.e. the total number of rows with NAs)
```{r}
sum(is.na(df$steps))
```
#### 2. Devise a strategy for filling in all of the missing values in the dataset. 
#### 3. Create a new dataset that is equal to the original dataset but with the missing data filled in.df_impute <- df
```{r}
df_impute <- df
ndx <- is.na(df_impute$steps)
int_avg <- tapply(df_ign$steps, df_ign$interval, mean, na.rm=TRUE, simplify=T)
df_impute$steps[ndx] <- int_avg[as.character(df_impute$interval[ndx])]
```
#### 4. Make a histogram of the total number of steps taken each day and Calculate and report the mean and median total number of steps taken per day.

```{r}
new_dailysum <- tapply(df_impute$steps, df_impute$date, sum, na.rm=TRUE, simplify=T)
hist(x=new_dailysum,
     col="red",
     breaks=20,
     xlab="daily steps",
     ylab="frequency",
     main="The distribution of daily total (with missing data imputed)")
	 
	 mean(new_dailysum)
	 median(new_dailysum)
```
----------------------------------------

## **Are there differences in activity patterns between weekdays and weekends?**

#### 1. Create a new factor variable in the dataset with two levels – “weekday” and “weekend” indicating whether a given date is a weekday or weekend day.
```{r}
is_weekday <- function(d) {
    wd <- weekdays(d)
    ifelse (wd == "Saturday" | wd == "Sunday", "weekend", "weekday")
}

wx <- sapply(df_impute$date, is_weekday)
df_impute$wk <- as.factor(wx)
head(df_impute)
```

#### 2. Make a panel plot containing a time series plot of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all weekday days or weekend days (y-axis). 
```{r}
wk_df <- aggregate(steps ~ wk+interval, data=df_impute, FUN=mean)

library(lattice)
xyplot(steps ~ interval | factor(wk),
       layout = c(1, 2),
       xlab="Interval",
       ylab="Number of steps",
       type="l",
       lty=1,
       data=wk_df)
```